Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2023-11-28 docs/Localized-Modeling/Localized-Modeling.Rmd
track second submission + redistribute bibliography files + script to copy figures

title: "Clinical prediction with localized modeling using similarity-based cohorts: A scoping review"
titlerunning: Clinical Prediction with Localized Modeling
authors: 
  - firstname: Adi
    lastname: Cohen
    degrees: BS?
    email: ac4569@my.nsu.edu
    affiliation: 1
  - firstname: Patti
    lastname: McCall-Junkin
    degrees: MA, MLS
    email: pattimccall.junkin@gmail.com
    affiliation: 2
  - firstname: Jason Cory
    lastname: Brunson
    degrees: PhD
    email: jason.brunson@medicine.ufl.edu
    orcid: 0000-0003-3126-9494
    affiliation: 3
    corresponding: TRUE
affiliations:
  - number: 1
    info:
      orgdiv: College of Osteopathic Medicine
      orgname: Nova Southeastern University
    address:
        street: 3200 S University Drive
        city: Davie
        postcode: 33328
        state: FL
        country: USA
  - number: 2
    info:
      orgdiv: Smathers Libraries Academic Research & Consulting Services
      orgname: University of Florida
    address:
        street: P.O. Box 117000
        city: Gainesville
        postcode: 32611-7000
        state: FL
        country: USA
  - number: 3
    corresponding: TRUE
    info:
      orgdiv: Laboratory for Systems Medicine
      orgname: University of Florida
    address:
        street: P.O. Box 100225 JHMHC
        city: Gainesville
        postcode: 32610-0225
        state: FL
        country: USA
keywords:
  - case-based reasoning
  - localized modeling
  - nearest neighbors
  - clinical decision support
  - scoping review
# pacs: 
#   jel:
#     - "D8"
#     - "H51"
#   msc:
#     - "35A01"
#     - "65L10"

@@

bibliography:
  - ../bibliography/SR-bibliography.bib
  - ../bibliography/SR-0-Found.bib
  - ../bibliography/SR-2-Included.bib
  - ../bibliography/SR-3-Tracked.bib

@@

### Acknowledgments {-}

We are grateful for the astute and supportive comments of three anonymous reviewers.

@@

Following the selection sources aligning with our review, we collected characteristics of included studies (Section \ref{data-and-code}). The features included bibliographic fields (date of publication, journal, authors, title, keywords, DOI), study goals (objective, generalizable knowledge, evaluation, clinical/medical domain), data sets (data source, type of data, range of data, number of cases/incidences, number of predictors/features), and methodological choices (types of similarity measure, families of adaptation step/statistical model, method(s) compared against, performance measures, results of evaluations and comparisons, name given to modeling approach). We used these data to detect and visualize trends amongst the included studies.

@@

```{r prisma, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "PRISMA-S flow chart.
#|  Lowercase letters refer to items obtained from
#|  the seed set (m),
#|  the structured search (n), and
#|  citation/reference tracking (s)."
knitr::include_graphics("../../fig/fig-prisma.pdf")
```

@@

The `r nrow(tab_synthesis)` included studies were analyzed based on several characteristics (see Section \ref{data-and-code}), and we report and describe some observations here (Table \ref{tab:synthesis}). The studies were published in a variety of journals, with some of greater frequency, though no single journal published more than 3. The journal _Evolving Systems_ published 2 of the included studies, _Artificial Intelligence in Medicine_ published 3, _Hindawi - Journal of Healthcare Engineering_ published 3, and _PLOS One_ published 2.

@@

@@

```{r year, fig.align='center', out.width="75%", echo=FALSE}
#| fig.cap =
#| "Number of publications in our sample each year."
knitr::include_graphics("../../fig/fig-years.png")
```

@@

@@

```{r properties, fig.align='center', out.width="100%", echo=FALSE}
#| fig.cap =
#| "Share of studies characterized by several design elements:
#|  data source (row),
#|  clinical task (fill color), and
#|  remit (opacity)."
knitr::include_graphics("../../fig/fig-properties.png")
```

## Rationales
\label{rationales}

The included studies hypothesized, asserted, or assumed several benefits of localized modeling specific to clinical and medical settings and advantages over other modeling approaches (see Section \ref{data-and-code}).
The most common was that the restriction to similar or relevant past cases would improve predictive performance for the index case [@Mariuzzi1997; @Liang2015a; @Ng2015; @Lee2017]. In particular, @Lee2015 hypothesized and confirmed that the value of each past case would be positively related to its similarity to the index case, an assumption built in to the weighting schemes of other approaches. @Lowsky2013 made a different case, that the fewer parametric assumptions and complications of a CBR-style model would allow for greater accuracy.
Additionally, several investigators asserted that the use of localized cohorts befit the clinical focus on the individual patient rather than the population, without reference to performance [@Song2006c; @Xu2008a].

@@

```{r methods, fig.align='center', out.width="100%", echo=FALSE}
#| fig.cap =
#| "Frequency of recurring methodological elements across included studies."
knitr::include_graphics("../../fig/fig-methods.png")
```

## Performance assessments
\label{performance-assessments}

@@

```{r framework, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "General framework for localized models.
#|  Dashed arrows indicate optional steps."
knitr::include_graphics("../../fig/fig-typology.pdf")
```

@@

# Support

The authors received no specific funding or other support for this work.

# Competing interests

The authors declare that they have no competing interests.

# Data and code

Search results are publicly available in a Zotero group library.
Data collected or encoded for included studies are publicly available in two Google Sheets.
Code used to analyze these data and generate this manuscript is publicly available in a GitHub repository.

* Search results, included studies, and other bibliography: <https://www.zotero.org/groups/5017571/imsr/>
* Bibliographic and methodological properties of included studies: <https://docs.google.com/spreadsheets/d/1tpWMhYH2pyRT55K7n2J2XFs-kEV_JTuCDmXzJ4BBgNo/>
* Terminology and composite techniques of included studies: <https://docs.google.com/spreadsheets/d/1xvDJwiLBoI2oz8fxHJ5MjNmiju_RAlK7RJv-wXe1DAs/>
* Code used to conduct analyses and prepare the manuscript: <https://github.com/corybrunson/imsr>

# Appendix

## Research question

@@

## Purpose of review

@@

## Search design

@@

### Search strings

@@

## Screening process

@@

## Selection process

@@

## Analysis and synthesis

@@

## Performance evaluations and comparisons

>>>>

title: "Clinical prediction with localized modeling using similarity-based cohorts: A scoping review"
titlerunning: Clinical Prediction with Localized Modeling
authors: 
  - firstname: Adi
    lastname: Cohen
    degrees: BS
    email: ac4569@my.nsu.edu
    affiliation: 1
  - firstname: Patti
    lastname: McCall-Junkin
    degrees: MA, MLS
    email: pattimccall.junkin@gmail.com
    affiliation: 2
  - firstname: Jason Cory
    lastname: Brunson
    degrees: PhD
    email: jason.brunson@medicine.ufl.edu
    orcid: 0000-0003-3126-9494
    affiliation: 3
    corresponding: TRUE
affiliations:
  - number: 1
    info:
      orgdiv: College of Osteopathic Medicine
      orgname: Nova Southeastern University
    address:
        street: 3200 S University Drive
        city: Davie
        postcode: 33328
        state: FL
        country: USA
  - number: 2
    info:
      orgdiv: Smathers Libraries Academic Research & Consulting Services
      orgname: University of Florida
    address:
        street: P.O. Box 117000
        city: Gainesville
        postcode: 32611-7000
        state: FL
        country: USA
  - number: 3
    corresponding: TRUE
    info:
      orgdiv: Laboratory for Systems Medicine
      orgname: University of Florida
    address:
        street: P.O. Box 100225 JHMHC
        city: Gainesville
        postcode: 32610-0225
        state: FL
        country: USA
keywords:
  - case-based reasoning
  - localized modeling
  - nearest neighbors
  - clinical decision support
  - scoping review
# pacs: 
#   jel:
#     - "D8"
#     - "H51"
#   msc:
#     - "35A01"
#     - "65L10"

@@

bibliography:
  - SR-bibliography.bib
  - SR-0-Found.bib
  - SR-2-Included.bib
  - SR-3-Tracked.bib

@@

\pagebreak

@@

Following the selection sources aligning with our review, we collected characteristics of included studies. The features included bibliographic fields (date of publication, journal, authors, title, keywords, DOI), study goals (objective, generalizable knowledge, evaluation, clinical/medical domain), data sets (data source, type of data, range of data, number of cases/incidences, number of predictors/features), and methodological choices (types of similarity measure, families of adaptation step/statistical model, method(s) compared against, performance measures, results of evaluations and comparisons, name given to modeling approach). We used these data to detect and visualize trends amongst the included studies.

@@

```{r prisma, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "PRISMA-S flow chart.
#|  Lowercase letters refer to items obtained from
#|  the seed set (m),
#|  the structured search (n), and
#|  citation/reference tracking (s)."
knitr::include_graphics("Fig1.pdf")
```

@@

The `r nrow(tab_synthesis)` included studies were analyzed based on several characteristics, and we report and describe some observations here (Table \ref{tab:synthesis}). The studies were published in a variety of journals, with some of greater frequency, though no single journal published more than 3. The journal _Evolving Systems_ published 2 of the included studies, _Artificial Intelligence in Medicine_ published 3, _Hindawi - Journal of Healthcare Engineering_ published 3, and _PLOS One_ published 2.

@@

\small

@@

\normalsize

```{r year, fig.align='center', out.width="100%", echo=FALSE}
#| fig.cap =
#| "Number of publications in our sample each year."
knitr::include_graphics("Fig2.png")
```

@@

\normalsize

@@

```{r properties, fig.align='center', out.width="100%", echo=FALSE}
#| fig.cap =
#| "Share of studies characterized by several design elements:
#|  data source (row),
#|  clinical task (fill), and
#|  remit (opacity)."
knitr::include_graphics("Fig3.png")
```

## Rationales

The included studies hypothesized, asserted, or assumed several benefits of localized modeling specific to clinical and medical settings and advantages over other modeling approaches.
The most common was that the restriction to similar or relevant past cases would improve predictive performance for the index case [@Mariuzzi1997; @Liang2015a; @Ng2015; @Lee2017]. In particular, @Lee2015 hypothesized and confirmed that the value of each past case would be positively related to its similarity to the index case, an assumption built in to the weighting schemes of other approaches. @Lowsky2013 made a different case, that the fewer parametric assumptions and complications of a CBR-style model would allow for greater accuracy.
Additionally, several investigators asserted that the use of localized cohorts befit the clinical focus on the individual patient rather than the population, without reference to performance [@Song2006c; @Xu2008a].

@@

```{r methods, fig.align='center', out.width="100%", echo=FALSE}
#| fig.cap =
#| "Frequency of recurring methodological elements across included studies."
knitr::include_graphics("Fig4.png")
```

## Performance assessments

@@

```{r framework, fig.align='center', out.width="100%", echo=FALSE}
#| fig.cap =
#| "General framework for localized models.
#|  Dashed arrows indicate optional steps."
knitr::include_graphics("Fig5.pdf")
```

@@

\pagebreak

# Appendix {-}

## Research question {-}

@@

## Purpose of review {-}

@@

## Search design {-}

@@

### Search strings {-}

@@

## Screening process {-}

@@

## Selection process {-}

@@

## Analysis and synthesis {-}

@@

## Performance evaluations and comparisons {-}

### 2023-11-28 docs/Manuscript/Manuscript.Rmd
track second submission + redistribute bibliography files + script to copy figures

---
title: "Clinical prediction with localized modeling using similarity-based cohorts: A scoping review"
author:
  - name: Adi Cohen
    email: ac4569@my.nsu.edu
    affiliation: 1
  - name: Patti McCall-Junkin
    email: pattimccall.junkin@gmail.com
    affiliation: 2
  - name: Jason Cory Brunson
    email: jason.brunson@medicine.ufl.edu
    affiliation: 3
    correspondingauthor: true
address:
  - code: 1
    address: College of Osteopathic Medicine, Nova Southeastern University, 3200 S University Drive, Davie, FL 33328
  - code: 2
    address: Smathers Libraries Academic Research & Consulting Services, University at Florida, P. O. Box 117000, Gainesville, FL 32611-7000
  - code: 3
    address: Laboratory for Systems Medicine, University of Florida, P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225
abstract: |
  Background:
  Applications of classical case-based reasoning (CBR) to clinical and health tasks have given rise to a family of techniques in which a statistical model is fitted to a subset of labeled cases, selected for relevance or similarity to an unlabeled case to which the model is applied. These localized models, as we propose to call them, may perform as well as "black-box" models while retaining direct interpretations. Their adoption has been hindered by fragmented development and implementation challenges. Our goal in this scoping review was to describe the clinical and health applications of localized models to date and to derive a general framework for their design and evaluation.

>>>>

---
title: "Clinical prediction with localized modeling using similarity-based cohorts: A scoping review"
author:
  - name: Adi Cohen
    email: ac4569@my.nsu.edu
    affiliation: 1
  - name: Patti McCall-Junkin
    email: pattimccall.junkin@gmail.com
    affiliation: 2
  - name: Jason Cory Brunson
    email: jason.brunson@medicine.ufl.edu
    affiliation: 3
    correspondingauthor: true
address:
  - code: 1
    address: College of Osteopathic Medicine, Nova Southeastern University, 3200 S University Drive, Davie, FL 33328
  - code: 2
    address: Smathers Libraries Academic Research & Consulting Services, University of Florida, P.O. Box 117000, Gainesville, FL 32611-7000
  - code: 3
    address: Laboratory for Systems Medicine, University of Florida, P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225
abstract: |
  Background:
  Applications of classical case-based reasoning (CBR) to clinical and health tasks have given rise to a family of techniques in which a statistical model is fitted to a subset of labeled cases, selected for relevance or similarity to an unlabeled case to which the model is applied. These localized models, as we propose to call them, may perform as well as "black-box" models while retaining direct interpretations. Their adoption has been hindered by fragmented development and implementation challenges. Our goal in this scoping review was to describe the clinical and health applications of localized models to date and to derive a general framework for their design and evaluation.
